---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/USCbiostats/partition.svg?branch=master)](https://travis-ci.org/USCbiostats/partition) [![Build status](https://ci.appveyor.com/api/projects/status/fofb9um8jqyyi6av?svg=true)](https://ci.appveyor.com/project/malcolmbarrett/partition-qaqc6)
 [![Coverage status](https://codecov.io/gh/USCbiostats/partition/branch/master/graph/badge.svg)](https://codecov.io/github/USCbiostats/partition?branch=master)
 [![CRAN status](https://www.r-pkg.org/badges/version-ago/partition)](https://cran.r-project.org/package=partition)
 <!-- badges: end -->
 
# partition

partition is a fast and flexible framework for agglomerative partitioning. partition uses an approach  called Direct-Measure-Reduce to create new variables that maintain the user-specified minimum level of information. Each reduced variable is also interpretable: the original variables map to one and only one variable in the reduced data set. partition is flexible, as well: how variables are selected to reduce, how information loss is measured, and the way data is reduced can all be customized.  

## Installation

You can install the partition from CRAN with:

``` r
install.packages("partition")
```

Or you can install the development version of partition GitHub with:

``` r
# install.packages("remotes")
remotes::install_github("USCbiostats/partition")
```

## Example

```{r}
library(partition)
set.seed(1234)
df <- simulate_block_data(c(3, 4, 5), lower_corr = .4, upper_corr = .6, n = 100)

#  don't accept reductions where information < .6
prt <- partition(df, threshold = .6)
prt

# return reduced data
partition_scores(prt)

# access mapping keys
mapping_key(prt)

unnest_mappings(prt)

# use a lower threshold of information loss
partition(df, threshold = .5, partitioner = part_kmeans())

# use a custom partitioner
part_icc_rowmeans <- replace_partitioner(
  part_icc, 
  reduce = as_reducer(rowMeans)
)
partition(df, threshold = .6, partitioner = part_icc_rowmeans) 
```

partition also supports a number of ways to visualize partitions and permutation tests; these functions all start with `plot_*()`. These functions all return ggplots and can thus be extended using ggplot2.

```{r, dpi = 320}
plot_stacked_area_clusters(df) +
  ggplot2::theme_minimal(14)
```

## Contributing 

Please read the [Contributor Guidelines](.github/CODE_OF_CONDUCT.md) prior to submitting a pull request to partition. Also note that this project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.


